# Case Study 3: COVID-19 Mortality in Chicago (March 2020 - March 2022) {#cook-county-covid}

## Introduction 

The Cook County Medical Examiner has made a dataset with COVID-19 related
deaths publicly available with the following intent: 

> Cook County Government has created the Medical Examiner COVID-19 Dashboard to provide direct, transparent access to critical information about COVID-19 deaths in the County for public health agencies, medical professionals, first responders, journalists, policymakers and residents. The Medical Examiner’s Office encourages visitors to use this data to explore trends, identify areas of concern and take appropriate action. The data can be utilized to identify communities that are most severely impacted by the virus and can inform proactive public policy.

Read more: https://datacatalog.cookcountyil.gov/stories/s/ttk4-trbu 

Data Source: https://datacatalog.cookcountyil.gov/Public-Safety/Medical-Examiner-Case-Archive-COVID-19-Related-Dea/3trz-enys

## Motivation, Research Questions, and Learning Objectives

This case study focuses on the following research questions: 

  - What are the gradients in COVID-19 mortality by race/ethnicity in relation
  to ABSMs measuring: racial/ethnic composition, the ICE for racialized economic
  segregation, percent of population living below the poverty line, percent of
  people living in crowded housing, and median income;
  
  - What are the differences in COVID-19 mortality rates by race/ethnicity 
  accounting for age?
  
  <!-- 
  TODO: 
    We have not yet done: 
      - Identification of the ABSM effects without regard for race/ethnicity 
      - We could still add more clear presentation of the racial/ethnic differences
        that don't incorporate ABSMs -- as it is, this is just shown in a map figure.
      - How the racial disparity varies across ZCTAs 
  --> 

In today's case example using these data, we'll show you how you can use the
`tidycensus` package to download relevant area based population estimates and
sociodemographic measures from the US Census American Community Survey (ACS) to
contextualize the deaths observed in this dataset.

We'll show you how we cleaned the data and merged in data from the ACS in the
**Cleaning the Data** section, but we're providing to you a cleaned dataset that
should allow you to pick up and follow along from the **Visualizing Your Data**
section onwards.

## Dependencies 


```{r} 
#| eval = FALSE,
#| warning = FALSE,
#| message = FALSE

# mission critical packages
library(tidycensus) 
library(tidyverse)
library(sf)
library(tigris)
library(mapview)
library(INLA)

# nice to have packages
library(magrittr) # for the %<>% and %$% pipe
library(janitor)
library(purrr)
library(Hmisc)
library(epitools)
```

## Cleaning the Data

```{r cleaning the data}
#| eval = FALSE,
#| echo = TRUE

# get population sizes from ACS -------------------------------------------
acs_vars <- tidycensus::load_variables(2019, dataset = 'acs5')

# the ACS 2015-2019 variables can be viewed online here: 
# https://api.census.gov/data/2019/acs/acs5/variables.html
# 
# the race/ethnicity and age-group stratified population estimates from ACS are
# in the B01001 table.
# 
# B01001_001 through _049 are the sex/gender overall population size estimates,
# and B01001A through B01001I are the race/ethnicity specific tables.  For the
# race/ethnicity specific tables the age-groups suffixes range from _001 to _031
# 
# in the following three steps, we programmatically construct the population
# size variables that we want to retrieve from the ACS since otherwise there are
# a lot of them to type out.
race_chars <-
  c(
    white = 'H',
    black = 'B',
    hispanic_or_latino = 'I'
  )

# for each of A through I, construct B01001*_001 through _031
sex_race_age_vars <-
  paste0(rep(paste0('B01001', race_chars, '_0'), each = 31),
         stringr::str_pad(
           1:31,
           width = 2,
           side = 'left',
           pad = '0'
         ))

# this adds on the sex/gender overall population estimates
sex_race_age_vars %<>% c(.,
                         paste0(
                           'B01001_0',
                           stringr::str_pad(
                             1:49,
                             width = 2,
                             side = 'left',
                             pad = '0'
                           )
                         ))

# split the label
acs_vars %<>%
  tidyr::separate(label,
           into = c('estimate', 'total', 'gender', 'age', 'subgroup'),
           sep = '!!')

# clean label values (remove unnecessary colon characters and leading/trailing spaces)
acs_vars %<>% mutate_at(.vars = vars(estimate, total, gender, age, subgroup),
                        ~ gsub(":", "", .) %>% stringr::str_trim())

# select only what we need
acs_vars %<>% select(name, total, gender, age, concept)

# get sex, race (incl. hispanic or latino), and age stratified population
# estimates
popsizes <-
  tidycensus::get_acs(
    geography = 'zip code tabulation area',
    state = 'IL',
    year = 2019,
    geometry = TRUE,
    variables = sex_race_age_vars,
    output = 'tidy'
  )

# join in our variable names
popsizes %<>% left_join(acs_vars, by = c('variable' = 'name'))

popsizes %<>% janitor::clean_names() # clean column names
popsizes %<>% select(-total) # remove total column that only contains 'total'

# take out the geometry for use later
zip_geometry <- popsizes %>% select(geoid) %>% unique()
# ggplot(zip_geometry) + geom_sf() # zip codes of illinois

# remove geometry before data reshaping
popsizes %<>% sf::st_drop_geometry()

if ('sf' %in% class(popsizes)) {
  stop("popsizes data has geometry, this can cause errors in the group_by and summarize steps using dplyr.")
}

# aggregate by sex/gender
popsizes %<>%
  group_by(geoid, name, age, concept) %>%
  summarize(
  estimate = sum(estimate),
  # here we're using the tidycensus::moe_sum function to create margin of error estimates for 
  # sums of population size estimates
  moe = tidycensus::moe_sum(moe = moe, estimate = estimate)
  )
```

Since the Cook County Medical Examiner Case Archive data includes records on 
deaths where the decedents' residential ZIP codes are inside as well as outside 
Cook County, Illinois, we want to restrict our dataset to deaths where their
county of residence was in Cook County. 

```{r get relevant zip codes}
#| echo = TRUE,
#| eval = FALSE

# identify relevant zip codes ---------------------------------------------

# we originally downloaded the county shapefiles using tigris, but 
# we found that those contained water area we wanted to remove. 
# 
# we tried using the tigris::erase_water function but it took a very 
# long time to run for us, so we found that the alternative was easier
# to implement:  downloading a county shapefile directly from the census 
# and using that one which came already with the water areas removed. 
# 
# We got our county shapefile for Cook county in what follows from here: 
# https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html 
counties <- read_sf('cb_2018_us_county_5m/cb_2018_us_county_5m.shp')

# get the map for cook county
cook_county <- counties %>% filter(
  # get the 5-digit FIPS or GEOID for Cook County programmatically by filtering
  # the tigris::fips_codes dataframe, or if you happen to know it's 17031 
  # you could code it explicitly instead
  GEOID == tigris::fips_codes %>% 
    filter(county == 'Cook County', state == 'IL') %$% 
    paste0(state_code, county_code))

# calculate the zip codes intersecting cook county
zips_with_overlap <- zip_geometry[st_intersects(zip_geometry, cook_county) %>% map_lgl( ~ length(.) > 0),]

# calculate each zip code total area
zips_with_overlap$total_area <- st_area(zips_with_overlap)

# calculate the intersection of cook county and each zip code tabulation area
zips_with_overlap_intersection <- st_intersection(zips_with_overlap, cook_county)
zips_with_overlap_intersection %<>% select(geoid)

# calculate the area of each intersection
zips_with_overlap_intersection$overlapping_area_w_cook_cty <- st_area(zips_with_overlap_intersection)

# drop geometry so we can merge this back in
zips_with_overlap_intersection %<>% sf::st_drop_geometry()

# add in overlap calculation
zips_with_overlap %<>% left_join(
  zips_with_overlap_intersection %>% select(geoid, overlapping_area_w_cook_cty),
  by = c('geoid')
)

# calculate the proportion of area for that zip code within cook county
zips_with_overlap %<>% mutate(
  prop_area_overlap = as.numeric(overlapping_area_w_cook_cty / total_area))

# filter for zip codes with 90% land area inside cook county
zips_with_over_90pct_overlap <- zips_with_overlap %>%
  filter(prop_area_overlap >= .9)

# Visualize Cook County and ZIPs with over 90 percent overlap
ggplot() + 
  geom_sf(data = cook_county, aes(color = "Cook County"), fill = 'white') + 
  geom_sf(
    data = zips_with_over_90pct_overlap,
    mapping = aes(color = "Zip Code Tabulation Areas"),
    size = .5,
    alpha = .6
  ) + 
  scale_color_manual(
    values = c("Cook County" = "#E69F00", "Zip Code Tabulation Areas" = "#56B4E9") 
  ) + 
  # theme_bw() + 
  guides(color = guide_legend(override.aes = list(fill = c('white', 'grey95')))) +
  scale_x_continuous(breaks = c(-88.3, -88.1, -87.9, -87.7, -87.5)) +   
  ggtitle("Cook County and Zip Code Tabulation Areas 90% Contained Within")

# save the plot
# ggsave("cook_county_and_zip_codes.png", width = 10, height = 6)

# group population sizes together by age groups and race/ethnicity and zip code
zips_with_population_estimates_by_race_ethnicity_and_age <-
  popsizes %>%
  filter(! is.na(age) & concept != 'SEX BY AGE') %>% # filter for overall population
  filter(
    geoid %in% zips_with_over_90pct_overlap$geoid) %$%
  left_join(zips_with_over_90pct_overlap, .)

# code the Census race/ethnicity groups
zips_with_population_estimates_by_race_ethnicity_and_age %<>% mutate(
  race_ethnicity = case_when(
    concept ==  "SEX BY AGE (BLACK OR AFRICAN AMERICAN ALONE)" ~ "Black, Hispanic or non-Hispanic",
    concept ==  "SEX BY AGE (HISPANIC OR LATINO)" ~ "Hispanic",
    concept ==  "SEX BY AGE (WHITE ALONE, NOT HISPANIC OR LATINO)" ~ "White, non-Hispanic"
  ))


# remove totals (across age groups) observations
zips_with_population_estimates_by_race_ethnicity_and_age %<>% filter(! is.na(age))
zips_with_population_estimates_by_race_ethnicity_and_age %<>% filter(! is.na(race_ethnicity))

# aggregate into larger age groups
zips_with_population_estimates_by_race_ethnicity_and_age %<>% mutate(
  age_group = case_when(
    age %in% c(
      "Under 5 years",
      "5 to 9 years",
      "10 to 14 years",
      "15 to 17 years",
      "18 and 19 years",
      "20 to 24 years"
    ) ~ "Under 25",
    age %in% c(
      "25 to 29 years",
      "30 to 34 years"
    ) ~ "25 to 34 years",
    age == "35 to 44 years" ~ "35 to 44 years",
    age == "45 to 54 years" ~ "45 to 54 years",
    age == "55 to 64 years" ~ "55 to 64 years",
    age == "65 to 74 years" ~ "65 to 74 years",
    age == "75 to 84 years" ~ "75 to 84 years",
    age == "85 years and over" ~ "85 years and over"
  ))

# sum age group population sizes by race_ethnicity
zips_with_population_estimates_by_race_ethnicity_and_age %<>% group_by(geoid, age_group, race_ethnicity) %>%
  summarize(
    estimate = sum(estimate, na.rm=T),
    moe = moe_sum(moe, estimate)
  )


```

Now we're ready to add the population size estimates for each race/ethnicity and
age-group by ZIP code.

```{r add population size estimates}
#| echo = TRUE,
#| eval = FALSE

# read in your data
cook_county_deaths <- readr::read_csv(
  "cook_county_mortality/Medical_Examiner_Case_Archive_-_COVID-19_Related_Deaths.csv")

# merge race/ethnicity and age-group specific denominators ----------------

# use janitor::clean_names to standardize column name syntax into snake_case
cook_county_deaths %<>% janitor::clean_names()

# code race/ethnicity into the following: 
#  Black (Hispanic or non-Hispanic)
#  Hispanic and non-Hispanic
#  White non-Hispanic
cook_county_deaths %<>%
  mutate(
    black = race == 'Black',
    white_nh = latino & race == 'White',
    hispanic = latino
  )

# we restrict our time-period to during March 2020 to March 2022 -- noting the 
# information on the dataset stories web page: 
# 
#  https://datacatalog.cookcountyil.gov/stories/s/ttk4-trbu
# 
#  Effective April 1, 2022, the Cook County Medical Examiner’s Office no longer
#  takes jurisdiction over hospital, nursing home or hospice COVID-19 deaths
#  unless there is another factor that falls within the Office’s jurisdiction.
# 
cook_county_deaths %<>% filter(
  lubridate::mdy_hms(date_of_death) >= lubridate::mdy("03/01/2020") & 
    lubridate::mdy_hms(date_of_death) <= lubridate::mdy("04/01/2022")
)


# Truncate Zip Codes to 5 Characters
cook_county_deaths %<>% mutate(residence_zip = stringr::str_extract(residence_zip, "^[0-9]{5}"))

# Categorize Age Group
cook_county_deaths %<>% mutate(
    age_group = case_when(
      age <= 24 ~ "Under 25",
      age >= 25 & age <= 35 ~ "25 to 34 years",
      age >= 35 & age <= 44 ~ "35 to 44 years",
      age >= 45 & age <= 54 ~ "45 to 54 years",
      age >= 55 & age <= 64 ~ "55 to 64 years",
      age >= 65 & age <= 74 ~ "65 to 74 years",
      age >= 75 & age <= 84 ~ "75 to 84 years",
      age >= 85 ~ "85 years and over",
      TRUE ~ NA_character_
    )
  )

cook_county_deaths$age_group %<>% factor(
  levels = c(
   "Under 25",
   "25 to 34 years",
   "30 to 34 years",
   "35 to 44 years",
   "45 to 54 years",
   "55 to 64 years",
   "65 to 74 years",
   "75 to 84 years",
   "85 years and over")
  )

# separate out deaths into the racial categories we're interested in 
# 
# note that these do have overlap: since the ACS variables do not include 
# age tables for the Black non-Hispanic group, we are using the Black 
# Hispanic or non-Hispanic population by age tables, and so there is overlap
# between them and the Hispanic group counts.
deaths_by_race_ethnicity_group <- 
  list(
    black = cook_county_deaths %>% filter(black),
    hispanic = cook_county_deaths %>% filter(hispanic),
    white_nh = cook_county_deaths %>% filter(white_nh)
  ) 

# Assign labels for the racialized groups 
race_ethnicity_groups <- c(
  "Black, Hispanic or non-Hispanic",
  "Hispanic",
  "White, non-Hispanic"
)

# for each racial group, tabulate the number of deaths by age group and 
# residence zip code. 
# 
# here, purrr::map is applying a function which does that tabulation to each of
# the data frames in the deaths_by_race_ethnicity_group list
deaths_by_race_ethnicity_group <- 
  purrr::map(1:3, function(i) {
    group_by(deaths_by_race_ethnicity_group[[i]], residence_zip, age_group) %>% 
        count(name = 'deaths') %>% 
      mutate(race_ethnicity = race_ethnicity_groups[[i]])
  })

# bind the tables for each of the race_ethnicity groups together
deaths_by_race_ethnicity_group %<>% bind_rows()

# check how many of the deaths proportionally were NA
ggplot(deaths_by_race_ethnicity_group, aes(x = age_group, y = deaths)) + 
  geom_col() + 
  facet_wrap(~race_ethnicity) + 
  ylab("Count of Deaths") + 
  theme(axis.text.x = element_text(angle = 75, hjust = 1))

# remove missing age, missing race/ethnicity deaths
deaths_by_race_ethnicity_group %<>% filter(! is.na(age_group))
deaths_by_race_ethnicity_group %<>% filter(! is.na(race_ethnicity))

# check distribution of deaths by age group
deaths_by_race_ethnicity_group %>% 
  group_by(age_group) %>% 
  count() 

# check distribution of deaths by racial/ethnic group
deaths_by_race_ethnicity_group %>% 
  group_by(race_ethnicity) %>% 
  count() 

# merge the denominators and deaths together
df <-
  zips_with_population_estimates_by_race_ethnicity_and_age %>% left_join(
  deaths_by_race_ethnicity_group,
  by = c('geoid' = 'residence_zip', 'age_group' = 'age_group', 'race_ethnicity' = 'race_ethnicity'))

# make age_group a factor variable
df$age_group %<>% factor(
  levels = c(
    'Under 25',
    '25 to 34 years',
    '35 to 44 years',
    '45 to 54 years',
    '55 to 64 years',
    '65 to 74 years',
    '75 to 84 years',
    '85 years and over'
  )
)

# add person-time
observation_time_in_years <- as.integer(lubridate::mdy("04-01-2022") - lubridate::mdy("03-01-2020")) / 365
df %<>% mutate(person_time = estimate * observation_time_in_years)
df %<>% mutate(deaths = ifelse(is.na(deaths), 0, deaths))
df %<>% mutate(mortality_per100k_py = deaths / person_time * 1e5)

df %>% 
  group_by(age_group, race_ethnicity) %>% 
  summarize(deaths = sum(deaths)) %>%
  ggplot(aes(x = age_group, y = deaths)) + 
  geom_col() + 
  facet_wrap(~race_ethnicity) + 
  xlab("Age Group") + 
  ylab("Count of Deaths") + 
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  ggtitle("Age Distribution of COVID-19 Deaths in Cook County IL")

df %>% 
  group_by(age_group, race_ethnicity) %>% 
  summarize(deaths = sum(deaths), 
            person_time = sum(person_time), 
            mortality_per100k_py = deaths / person_time * 1e5,
            mortality_ci_high = (1e5/person_time) * (deaths + sqrt(deaths)*1.96),
            mortality_ci_low = (1e5/person_time) * (deaths - sqrt(deaths)*1.96)
            ) %>%
  ggplot(
    aes(
      x = age_group,
      y = mortality_per100k_py,
      ymax = mortality_ci_high,
      ymin = mortality_ci_low,
      color = race_ethnicity,
      shape = race_ethnicity
    )
  ) + 
  geom_pointrange(position = position_dodge(width=.4), width = .2) + 
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_log10() + 
  xlab("Age Group") + 
  ylab("Mortality Rate per 100,000 Person Years") + 
  labs(color = 'Race/Ethnicity', shape = 'Race/Ethnicity') + 
  ggtitle("COVID-19 Mortality Rate by Age and Race/Ethnicity in Cook County IL")
```

We're in shape to add our area based socioeconomic metrics. We'll add the
Index of Concentration at the Extremes for Racialized Economic Segregation, the
proportion of the population under the poverty line, and the median income in
each ZIP code.

```{r add absms}
#| echo = TRUE,
#| eval = FALSE


# add area based socioeconomic measures  ----------------------------------

# get zip code rates for
#   - poverty
#   - ICEraceinc
#   - median income

absms_dictionary <- tibble::tribble(
  ~var, ~varname, ~description,
  # total population
  "B01001_001",  "total_popsize", "total population estimate", 
  
  # racial composition 
  'B01003_001',  "race_ethnicity_total", "race_ethnicity_total",
  
  # ICEraceinc
  "B19001_001",  'hhinc_total',   "total population for household income estimates",
  "B19001A_002", 'hhinc_w_1',     "white n.h. pop with household income <$10k",
  "B19001A_003", 'hhinc_w_2',     "white n.h. pop with household income $10k-14 999k",
  "B19001A_004", 'hhinc_w_3',     "white n.h. pop with household income $15k-19 999k",
  "B19001A_005", 'hhinc_w_4',     "white n.h. pop with household income $20k-24 999k",
  "B19001A_014", 'hhinc_w_5',     "white n.h. pop with household income $100 000 to $124 999",
  "B19001A_015", 'hhinc_w_6',     "white n.h. pop with household income $125k-149 999k",
  "B19001A_016", 'hhinc_w_7',     "white n.h. pop with household income $150k-199 999k",
  "B19001A_017", 'hhinc_w_8',     "white n.h. pop with household income $196k+",
  "B19001_002",  'hhinc_total_1', "total pop with household income <$10k",
  "B19001_003",  'hhinc_total_2', "total pop with household income $10k-14 999k",
  "B19001_004",  'hhinc_total_3', "total pop with household income $15k-19 999k",
  "B19001_005",  'hhinc_total_4', "total pop with household income $20k-24 999k",

  # poverty
  "B05010_002",  'in_poverty',    "population with household income < poverty line",
  "B05010_001",  'total_pop_for_poverty_estimates',  "total population for poverty estimates",

  # median income
  "B06011_001",  'median_income',  "median income estimate for total population",
  
  # crowded housing
  "B25014_005",  'owner_occupied_crowding1', 'owner occupied, 1 to 1.5 per room',
  "B25014_006",  'owner_occupied_crowding2', 'owner occupied, 1.51 to 2 per room',
  "B25014_007",  'owner_occupied_crowding3', 'owner occupied, 2.01 or more per room',
  "B25014_011",  'renter_occupied_crowding1', 'owner occupied, 1 to 1.5 per room',
  "B25014_012",  'renter_occupied_crowding2', 'owner occupied, 1.51 to 2 per room',
  "B25014_013",  'renter_occupied_crowding3', 'owner occupied, 2.01 or more per room',
  "B25014_001",  'crowding_total',            'total for crowding (occupants per room)',
  
  "B01001I_001",  'total_hispanic',           'total hispanic population estimate',
  "B01001B_001",  'total_black',              'total black, hispanic or non-hispanic estimate',
  "B01001H_001",  'total_white_nh',           'total white, non-hispanic population estimate'
)

get_absms <- function(zip_codes) {
  absms <- tidycensus::get_acs(
    year = 2019,
    geography = 'zcta',
    state = 'IL',
    zcta = zip_codes,
    variables = absms_dictionary$var,
    geometry = FALSE
  )

  # pivot wider
  absms %<>% select(-moe) %>%
    tidyr::pivot_wider(names_from = variable, values_from = estimate)

  rename_vars <- setNames(absms_dictionary$var, absms_dictionary$varname)
  absms <- absms %>% rename(!!rename_vars)

  absms %<>%
    mutate(
      # we calculate the people of color low income counts as the overall
      # low income counts minus the white non-hispanic low income counts
      people_of_color_low_income =
        (hhinc_total_1 + hhinc_total_2 + hhinc_total_3 + hhinc_total_4) -
        (hhinc_w_1 + hhinc_w_2 + hhinc_w_3 + hhinc_w_4),
      # sum up the white non-hispanic high income counts
      white_non_hispanic_high_income =
        (hhinc_w_5 + hhinc_w_6 + hhinc_w_7 + hhinc_w_8),
      # calculate the index of concentration at the extremes for racialized
      # economic segregation (high income white non-hispanic vs. low income
      # people of color)
      ICEraceinc =
        (white_non_hispanic_high_income - people_of_color_low_income) /
        hhinc_total,

      prop_in_poverty = in_poverty / total_pop_for_poverty_estimates,
      
      crowding = (owner_occupied_crowding1 + owner_occupied_crowding2 + owner_occupied_crowding3 +
        renter_occupied_crowding1 + renter_occupied_crowding2 + renter_occupied_crowding3) / crowding_total,
      
      prop_black = total_black / total_popsize,
      prop_hispanic = total_hispanic / total_popsize,
      prop_white_nh = total_white_nh / total_popsize
    ) %>%
    select(GEOID, ICEraceinc, prop_in_poverty, median_income, crowding, prop_black, prop_hispanic, prop_white_nh)

  return(absms)
}

absms <- get_absms(unique(df$geoid))

# merge in data
df %<>% left_join(absms, by = c('geoid' = 'GEOID'))

# get cutpoints for ICEraceinc in illinois --------------------------------

# we're doing this because we want our cutpoints to be in reference to the 
# state distribution for the ICEraceinc variable;  other cutpoints can be used, 
# but it's important to be transparent about what cutpoints are used and consider
# how the choice of cutpoints may affect the analysis.
# 
illinois_absms <-
  tidycensus::get_acs(
    year = 2019,
    geography = 'zcta',
    state = 'IL',
    variables = absms_dictionary$var,
    geometry = FALSE
  ) %>%
  select(-moe) %>%
  tidyr::pivot_wider(names_from = variable, values_from = estimate)

rename_vars <- setNames(absms_dictionary$var, absms_dictionary$varname)
illinois_absms %<>% rename(!!rename_vars)

illinois_absms %<>%
  mutate(
    # we calculate the people of color low income counts as the overall
    # low income counts minus the white non-hispanic low income counts
    people_of_color_low_income =
      (hhinc_total_1 + hhinc_total_2 + hhinc_total_3 + hhinc_total_4) -
      (hhinc_w_1 + hhinc_w_2 + hhinc_w_3 + hhinc_w_4),
    # sum up the white non-hispanic high income counts
    white_non_hispanic_high_income =
      (hhinc_w_5 + hhinc_w_6 + hhinc_w_7 + hhinc_w_8),
    # calculate the index of concentration at the extremes for racialized
    # economic segregation (high income white non-hispanic vs. low income
    # people of color)
    ICEraceinc =
      (white_non_hispanic_high_income - people_of_color_low_income) /
      hhinc_total,
    
    # we don't need poverty here since we're going to use pre-specified cutpoints
    # of 0-5%, 5-10%, 10-20% and 20%+ 
    
    # calculate crowding
    crowding = (owner_occupied_crowding1 + owner_occupied_crowding2 + owner_occupied_crowding3 +
        renter_occupied_crowding1 + renter_occupied_crowding2 + renter_occupied_crowding3) / crowding_total,
    
    # racial/ethnic composition variables
    prop_black = total_black / total_popsize,
    prop_hispanic = total_hispanic / total_popsize,
    prop_white_nh = total_white_nh / total_popsize
  ) 

illinois_ICEraceinc_cutpoints <-
  Hmisc::wtd.quantile(illinois_absms$ICEraceinc,
                      illinois_absms$total_popsize,
                      seq(0, 1, .2),
                      na.rm = T)

illinois_median_income_cutpoints <-
  Hmisc::wtd.quantile(illinois_absms$median_income,
                      illinois_absms$total_popsize,
                      seq(0, 1, .2),
                      na.rm = T)

illinois_crowding_cutpoints <-
  Hmisc::wtd.quantile(illinois_absms$crowding,
                      illinois_absms$total_popsize,
                      seq(0, 1, .2),
                      na.rm = T)

illinois_prop_hispanic_cutpoints <-
  Hmisc::wtd.quantile(illinois_absms$prop_hispanic,
                      illinois_absms$total_popsize,
                      seq(0, 1, .2),
                      na.rm = T)

illinois_prop_black_cutpoints <-
  Hmisc::wtd.quantile(illinois_absms$prop_black,
                      illinois_absms$total_popsize,
                      seq(0, 1, .2),
                      na.rm = T)

illinois_prop_white_nh_cutpoints <-
  Hmisc::wtd.quantile(illinois_absms$prop_white_nh,
                      illinois_absms$total_popsize,
                      seq(0, 1, .2),
                      na.rm = T)

# create cutpoint-leveled version of key ABSM variables
df$ICEraceinc_cut <- df$ICEraceinc %>% cut(., illinois_ICEraceinc_cutpoints, include.lowest=TRUE)
df$median_income_cut <- df$median_income %>% cut(., illinois_median_income_cutpoints, include.lowest=TRUE)
df$crowding_cut <- df$crowding %>% cut(., illinois_crowding_cutpoints, include.lowest=TRUE)
df$prop_black_cut <- df$prop_black %>% cut(., illinois_prop_black_cutpoints, include.lowest=TRUE)
df$prop_hispanic_cut <- df$prop_hispanic %>% cut(., illinois_prop_hispanic_cutpoints, include.lowest=TRUE)
df$prop_white_nh_cut <- df$prop_white_nh %>% cut(., illinois_prop_white_nh_cutpoints, include.lowest=TRUE)
```

We have a few remaining necessary cleaning steps to perform before we're ready
to model these data. Firstly, we're going to add discretized (cut) versions of
the percent in poverty and median income measures.  This is because this helps
to allow the model flexibility to fit nonlinear responses with increasing levels
in these covariates.  Alternative approaches could involve fitting the models
with smoothing splines on these variables instead, but we aren't showing that
here today.

Additionally, we have to make sure our factor variables have appropriate 
reference levels set.  Typically we set the reference level as the most 
privileged group so that we can frame the results as "the _____ group has
X times the mortality rate of the reference group." 

```{r final cleaning steps}
#| echo = TRUE,
#| eval = FALSE
# prepare data for modeling  ----------------------------------------------

# remove infinite or NA mortality rates since they will cause errors in trying
# to fit the models
df_prepped <- df %>% filter(
  is.finite(mortality_per100k_py) &
    ! is.na(mortality_per100k_py))

# ungroup
df_prepped %<>% ungroup()

# make the most privileged the reference category
df_prepped %<>% mutate(
  race_ethnicity = forcats::fct_relevel(factor(race_ethnicity), "White, non-Hispanic"),
  age_group = age_group,
  ICEraceinc_cut = forcats::fct_rev(ICEraceinc_cut)
)

# add median income zscore since otherwise the model coefficients on income
# are very small
zscore <- function(x) { (x - mean(x, na.rm=T)) / sd(x, na.rm=T) }
df_prepped$median_income_zscore <- zscore(df_prepped$median_income)

# add in cutpoints for poverty and median income quantiles
df_prepped %<>%
  ungroup() %>%
  mutate(
  prop_in_poverty_cut = cut(prop_in_poverty, c(0, .05, .1, .2, 1), include.lowest=T),
  median_income_quantiles = cut(median_income, quantile(median_income, seq(0,1,.2), na.rm=T), include.lowest=T)
)

# rename the estimate variable to 'population_estimate' to be more clear
df_prepped %<>% rename(population_estimate = estimate)

# save data 
write.csv(df_prepped, "cook_county_mortality_cleaned.csv", row.names=F)
saveRDS(df_prepped, "cook_county_mortality_cleaned.rds")
```

## Visualizing Your Data

```{r visualize data, fig.width = 8, fig.height = 8}
#| echo = TRUE,
#| eval = FALSE
df <- readRDS("data/09-cook-county-covid/cook_county_mortality_cleaned.rds")

# let's start by mapping the crude mortality rates in each race/ethnicity and 
# age strata

df %<>% mutate(deaths = ifelse(is.na(deaths), 0, deaths))
df %<>% mutate(mortality_per100k_py = deaths / person_time * 1e5)

# plot raw mortality rates
df %>% 
  ggplot(aes(fill = mortality_per100k_py)) + 
  geom_sf(data = cook_county, fill = 'dimgrey') + 
  geom_sf(size = 0) + 
  facet_grid(forcats::fct_rev(race_ethnicity)~age_group) + 
  scale_fill_distiller(palette = 'Reds', trans = scales::pseudo_log_trans(sigma = 100), direction = 1, 
                       labels = scales::comma_format(), na.value = 'dimgrey',
                       breaks = c(0, 1000, 10000, 80000)) +
  scale_x_continuous(breaks = c( -88.1, -87.7)) +   
  theme(legend.position = 'bottom',
        legend.text = element_text(angle = 75, hjust=1)
        ) + 
  ggtitle("Crude Mortality Rates by ZCTA, Race/Ethnicity and Age Group")

ggsave("raw_rates_by_zcta.png", width = 14, height = 8)

df %>% ggplot(aes(x = mortality_per100k_py, fill = race_ethnicity)) +
  geom_histogram(alpha = .8, position = 'identity') +
  facet_grid(age_group ~ forcats::fct_rev(race_ethnicity),
             scales = 'free') +
  scale_x_continuous(
    trans = scales::pseudo_log_trans(sigma = 10),
    labels = scales::comma_format(),
    breaks = c(0, 100, 1000, 5e4)
  ) + 
  xlab("Mortality Rate per 100,000 Person Years") + 
  theme(legend.position = 'bottom')

ggsave("raw_rates_histogram.png",
       width = 8,
       height = 16)

ggsave("raw_rates_histogram.png",
       width = 8,
       height = 16)

df %>% ggplot(aes(x = population_estimate, y = mortality_per100k_py, color = race_ethnicity)) +
  geom_point(alpha = .8) +
  facet_grid(age_group ~ forcats::fct_rev(race_ethnicity)) +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 10),
    labels = scales::comma_format(),
    breaks = c(0, 100, 1000, 5e4)
  ) + 
  scale_x_continuous(
    trans = scales::log_trans(),
    labels = scales::comma_format(),
  ) + 
  ylab("Mortality Rate per 100,000 Person Years") + 
  xlab("Population Size") + 
  theme(legend.position = 'bottom')

ggsave("raw_rates_scatter.png",
       width = 8,
       height = 16)


# plot population sizes
df %>% ggplot(aes(fill = population_estimate)) + 
  geom_sf(data = cook_county, fill = 'dimgrey') + 
  geom_sf(size = 0) + 
  facet_grid(forcats::fct_rev(race_ethnicity)~age_group) + 
  scale_fill_distiller(palette = 'Greens', trans = scales::pseudo_log_trans(sigma = 100), direction = 1, 
                       labels = scales::comma_format(), na.value = 'dimgrey',
                       limits = c(0, NA), 
                       breaks = c(0, 1000, 10000, 30000)) +
  scale_x_continuous(breaks = c( -88.1, -87.7)) +   
  theme(legend.position = 'bottom',
        legend.text = element_text(angle = 75, hjust=1)
        ) + 
  labs(fill = 'Population Estimate') + 
  ggtitle("Population Size by ZCTA, Race/Ethnicity and Age Group")

ggsave("population_size_by_zcta.png", width = 14, height = 8)

```

```{r visualize racial/ethnic breakdown by zip, fig.width = 8, fig.height = 4}
#| echo = TRUE,
#| eval = FALSE

# in the following visualizations the motivating principle behind the direction of the 
# color palette is to show (for sequential palettes) higher density with darker colors; 
# for the ICEraceinc variable a divergent color palette is used to draw attention to the 
# extreme ends of the scale

# visualize the proportional racial/ethnic breakdown
df %>% select(geoid, prop_black, prop_white_nh, prop_hispanic) %>% 
  tidyr::pivot_longer(
    cols = c(prop_black, prop_white_nh, prop_hispanic),
    names_to = "race_ethnicity",
    values_to = 'proportion'
  ) %>% 
  mutate(
    race_ethnicity = recode(race_ethnicity, 
                            prop_white_nh = 'White, non-Hispanic',
                            prop_black = 'Black, Hispanic or non-Hispanic',
                            prop_hispanic = 'Hispanic',
                            )
  ) %>% 
  ggplot(aes(fill = proportion)) + 
  geom_sf(size = 0) + 
  facet_grid(~race_ethnicity) + 
  scale_fill_distiller(
    palette = "Greens",
    direction = 1,
    labels = scales::percent_format()
  ) + 
  theme_dark() + 
  labs(fill = "Population Percentage") + 
  ggtitle("Racial/Ethnic Composition of ZCTAs in Cook County") + 
  theme(legend.position = 'bottom',
        legend.text = element_text(angle = 75, hjust=1),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
        )

ggsave("racial_ethnic_composition.png", width = 12, height = 8)

# visualize median income
df %>% 
  ggplot(aes(fill = median_income)) + 
  geom_sf(size = 0) + 
  scale_fill_viridis_c(direction = 1, 
                       labels = scales::dollar_format()) +
  theme_dark() + 
  labs(fill = 'Median Income') + 
  theme(legend.position = 'bottom',
        legend.text = element_text(angle = 75, hjust=1),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
        ) + 
  ggtitle("Median Income in ZCTAs in Cook County")

ggsave("median_income.png", width = 8, height = 8)

# visualize ICEraceinc
df %>% 
  ggplot(aes(fill = ICEraceinc)) + 
  geom_sf(size = 0) + 
  scale_fill_distiller(palette = 'BrBG', direction = 1, limits = c(-1,1),
                       labels = scales::number_format(accuracy = .01)) +
  theme_dark() + 
  labs(fill = paste0("High Income (>$100k annual household income) White non-Hispanic (high) vs.\n",
                     "Low Income (<$25k annual household income) People of Color (low)")) + 
  ggtitle("Index of Concentration at the Extremes for Racialized Economic Segregation",
          "ZCTAs in Cook County") + 
  guides(fill = guide_colourbar(
    title.position = "top",
    title.hjust = 0.5,
    barwidth = 10,
    
  ))+
  theme(
    legend.position = 'bottom',
        legend.title = element_text(size = 9),
        legend.text = element_text(angle = 75, hjust = 1),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
        )

ggsave("ICEraceinc.png", width = 8, height = 8)


# visualize crowding 
df %>% 
  ggplot(aes(fill = crowding)) + 
  geom_sf(size = 0) + 
  theme_dark() + 
    scale_fill_viridis_c(direction = 1, 
                       labels = scales::percent_format()) +
  labs(fill = "% Household Crowding") + 
  ggtitle("Household Crowding",
          "ZCTAs in Cook County") + 
  theme(
    legend.position = 'bottom',
        legend.title = element_text(size = 9),
        legend.text = element_text(angle = 75, hjust = 1),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
        )

ggsave("crowding.png", width = 8, height = 8)


# visualize poverty
df %>% 
  ggplot(aes(fill = prop_in_poverty)) + 
  geom_sf(size = 0) + 
  scale_fill_distiller(palette = 'Purples', direction = 1, limits = c(0,NA),
                       labels = scales::percent_format()) +
  theme_dark() + 
  labs(fill = 'Poverty Level') + 
  ggtitle("Poverty Level, ZCTAs in Cook County") + 
  theme(legend.position = 'bottom',
        legend.text = element_text(angle = 75, hjust=1),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
        )

ggsave("poverty.png", width = 8, height = 8)
```

