## Cleaning Area Identifiers

One of the most important aspects to pay attention to when cleaning
georeferenced data is that area identifiers are stored as
character or factor data types and not numeric types because if they are stored
as numeric types leading 0s will be dropped and this may cause issues when 
merging multiple datasets together if the area-keys are not coded in the same 
way (e.g. in one dataset area-keys might be coded numeric and in another dataset 
area-keys might be coded as a character or factor). 

This can happen often with FIPS codes, like the 2-character state FIPS
codes and 3-digit county FIPS code. For states like Alabama and Alaska with FIPS
codes `01` and `02`, if these are mistakenly stored as numeric values, they will
be truncated to as `1` and `2` which can introduce errors when trying to merge
multiple datasets using FIPS codes.

:::: {.infobox .dataWrangling}
For county FIPS codes, the `tigris` package has a handy built-in reference. 

Once you've installed `tigris` (e.g. run `install.packages('tigris')`) and
load the package (`library(tigris)`) you can access the built-in `fips_codes`
data.frame.

```{r}
#| echo = FALSE,
#| message = FALSE,
#| eval = TRUE
library(knitr)
library(kableExtra)
knitr::kable(tigris::fips_codes) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::scroll_box(height = '500px')
```
::::

Suppose, as can happen, that you download your data and find that due to a
coding error the 5-digit combined state and county FIPS codes have been stored
as numeric, causing leading zeroes to be truncated off.  If you check the FIPS
codes and are reasonably confident that the only error is that leading
left-hand-side zeroes have been omitted, you could do the following to correct
the mistake:

```{r}
#| echo = FALSE,
#| eval = TRUE
example_df <- tibble::tribble(
    ~FIPS,      ~Name, ~State,
  "01001",  "Autauga",   "AL",
  "01003",  "Baldwin",   "AL",
  "01005",  "Barbour",   "AL",
  "01007",     "Bibb",   "AL",
  "01009",   "Blount",   "AL",
  "01011",  "Bullock",   "AL",
  "01013",   "Butler",   "AL",
  "01015",  "Calhoun",   "AL",
  "01017", "Chambers",   "AL",
  "01019", "Cherokee",   "AL",
  "01021",  "Chilton",   "AL"
  )
example_df$FIPS <- as.integer(example_df$FIPS)
library(dplyr)
library(stringr)
```

```{r}
#| echo = TRUE,
#| eval = TRUE
head(example_df)

library(stringr)
library(dplyr)
example_df <- example_df %>% mutate(FIPS = ifelse(
  nchar(FIPS) == 4,
  str_pad(
    FIPS,
    width = 5,
    pad = '0',
    side = 'left'
  ),
  FIPS
))

head(example_df)
```

The above code uses the `dplyr` and `stringr` packages, both part of the `tidyverse`
and which have introductions here: https://dplyr.tidyverse.org/ and here: https://stringr.tidyverse.org/. 
