## Data Sources 

Health data can be found in numerous formats, such as for download from a
webpage as .csv or .xlsx files, accessible through querying an application
programming interface (API), or available only through direct request to a
health department or agency.

The sources for the data we've used in our case examples are as follows: 

  - US Census and American Community Survey data are retrieved through their API https://www.census.gov/data/developers/data-sets.html via the `tidycensus`
  package
  - CDC Places data were downloaded directly as .csv files from https://www.cdc.gov/places/
  - Cook County Medical Case Examiner Archive data on COVID-19 deaths were 
  downloaded directly from https://datacatalog.cookcountyil.gov/Public-Safety/Medical-Examiner-Case-Archive-COVID-19-Related-Dea/3trz-enys 
  - Massachusetts mortality data were requested directly from the Massachusetts 
  Department of Public Health
  - The Social Vulnerability Index (SVI) is available online to download with csv and 
  shapefiles available from https://www.atsdr.cdc.gov/placeandhealth/svi/index.html

## Loading Spreadsheet Data into R

If you have a .csv file that you have downloaded, you can use the `readr` package
(which is part of the `tidyverse`) to load it into R.

```{r}
#| echo = TRUE,
#| eval = FALSE
library(readr)
example_df <- read_csv("filename.csv")

# learn more about the options read_csv has by running:
?read_csv
```

Health data often come in other kinds of delimited formats such as 
tab-delimited or fixed-width spaced, in which case you can use the 
`read_tsv` or `read_fwf` functions from the `readr` package similarly 
to how you would use `read_csv`. Learn more about `readr` here: https://readr.tidyverse.org/

To read Excel data, we recommend using the `readxl` package, also part of 
the `tidyverse`. Learn more here: https://readxl.tidyverse.org/

```{r}
#| echo = TRUE,
#| eval = FALSE
library(readxl)
example_df <- read_excel("filename.xlsx", sheet = 1)
# to learn more about the options in the read_excel function, run:
?read_excel

# one particularly helpful feature to know about is the range argument which
# allows the user to specify they want to read a dataframe from a specific 
# range of cells using Excel-style range syntax:
example_df <- read_excel("filename.xlsx", sheet = 1, range = "A3:C17")
```

Even if your data are not a delimited text document or an Excel file, if they
are in a common file format, it is still quite likely you can read your data
into R using other packages. For example the `haven` package allows users to
read SAS, SPSS, and Stata files. Read more here <https://haven.tidyverse.org/>

## Connecting to Databases

Many online health datasets are accessible via query to a remote database
server.  References on how to interact with databases in R are available here:
https://db.rstudio.com/, and the following reference shows how to interact with
a remote database in the `tidyverse` style: https://dbplyr.tidyverse.org/
